A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-06-12 example.tex
Use a new, unsatisfiable example
This version uses a new example which is hopefully easier to follow. The linear algebra is slightly breakneck, but there is little I can do about that.

This section is now, hopefully, less jargon-heavy!

Closes: #7, #31, #26, #30,  #29, #28.

\begin{figure}[h]
    \centering 
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{a}
    \caption{The automaton $A'$, whose Parikh registers count the length of the
    string accepted by the automaton (register $l_A$), the start offset of the
    substring ($i$) and length ($n$) of the substring matching
    $\Regex{c\RegexOr{}dd}$.}\label{fig:aut_a}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{b}
    \caption{The automaton $B$, where the Parikh registers count the number of
    occurrences of each letter.}\label{fig:aut_b}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The collection of automata we use as running
  examples}\label{fig:examples}
\end{figure}

In this section we will introduce an intuition for how a string constraint
problem in an automata-based solver like \OstrichPlus{} is translated to a
Parikh automata intersection problem and solved efficiently using \Calculus{}.
First, by a Parikh automaton we mean a standard non-deterministic finite
automaton (NFA) with a set of Parikh registers that are incremented (or
decremented) at each transition. In the automata of \cref{fig:examples}, we will
use the notation $a / \begin{bmatrix} x_+ \end{bmatrix}$ to mean that a
transition reads an input character $a$ and increments register $x$ by one. We
will omit any zero-valued increments and assume that all registers are scoped to
their automaton. The increments are usually represented as a vector (hence the
notation), but as it is mostly sparse here, we instead use the symbolic notation
$\begin{bmatrix} x_{1+} \end{bmatrix}$ rather than the more cumbersome $\left[
0, 1,  0 , \ldots, 0 \right]$.

Furthermore, we use $\CountOf{c}(A)$ to refer to the number of letters $c$ in
$\Language(A)$, and allow this expression to be used both in linear constraints
of the input language of \cref{ex:string-constraints} and in the langauge of
constraints that we translate to. Where the automaton $A$ is apparent, we will
omit it.

We use standard PCRE regular expressions here and throughout the paper, writing
them $\Regex{\texttt{like this}}$. This means that $\RegexOr{}$ is alternation,
$\mathtt{*}$ the Kleene star, and $\mathtt{.}$, matches any single character.

\begin{example}\label{ex:string-constraints} Consider the following set of
    string constraints:
\begin{constraints}
    \item\label{const:x-in-c-dd} $x \in \Language(\Regex{c\RegexOr{}dd})$.
    \item\label{const:y-in-b} $y \in \Language(B)$, the language accepted by
    automaton~$B$ of \cref{fig:aut_b}.
    \item\label{const:x-substring} $x = \Substring(y, i, n)$, that is $x$ is an
    $n$-length substring of $y$ starting at offset $i$.
    \item\label{const:more-a-than-b} $i = n$, that is the substring is as long as its starting offset.
\end{constraints}
\end{example}

Intuitively, we see that the constraints are satisfiable.
\cref{const:more-a-than-b} implies that the starting offset must be either $1$
or $2$, that is we need to fill out $y$ with either one or two symbols before
the substring $x$ starts. The state $D$ is dead, as $B$ contains no transitions
with label $\Char{d}$, and hence none of the transitions into $D$ can appear in
the intersection of the languages. An example satisfying assignment is $y =
\Char{acca}$, which gives $x = \Char{c}$, $i = 1$, $n = 1$, but others are
possible.

To solve \cref{ex:string-constraints} using \Calculus{},
\cref{const:x-in-c-dd,const:y-in-b,const:x-substring} can be translated to the
product of Parikh automata $B \times A'$, where $A'$ (\cref{fig:aut_a}) is the
Parikh automata obtained by applying the encoding of $\Substring$ described in
\cite{ostrich-plus}. That description can then be used to obtain values for the
subsstring offsets $i, n$. Here follows an intuition for our method on $B \times
A'$ under the global constraint on the substring \cref{const:more-a-than-b}.

Our approach to the computation of the Parikh image (and therefore the solution
to the intersection problem posed by \cref{ex:string-constraints}) consists of
two applications of laziness; late computation of products of automata, and lazy
enforcement of connectivity constraints. Both are in contrast to the eager
approach to finding the Parikh image described in~\cite{generate-parikh-image},
that would have started by computing the product $A' \times B$, translated it to
a Presburger formula with $i, n$ etc as free variables, and then plugged in the
constraint \cref{const:more-a-than-b}.

Starting with an interactive theorem prover and the goal of solving the
constraint $\CountOf{\Char{a}}(B \times A') > \CountOf{\Char{b}}(B \times A')$
we associate each transition of $A'$ and $B$ with a fresh integer variable,
$\TransitionVar_\Transition$, identically to \cite{generate-parikh-image}. These
variables represent how many times each transition is taken. Hence, the final
counter value for each automaton is the element-wise sum:
\begin{equation}
\begin{bmatrix} 
  \CountOf{\Char{a}} \\ \CountOf{\Char{b}} 
  \\ \CountOf{\Char{c}} \\ \CountOf{\Char{d}} 
\end{bmatrix} = \sum_\Transition \TransitionVar_\Transition \cdot 
  \IncrementVec_\Transition \text{ where $\IncrementVec_\Transition$ are the increments of transition $\Transition$}.
\end{equation}

@@

We proceed by adding linear constraints requiring transition variables to
represent a flow through either automata by requiring that the number of
incoming transitions is equal to the number of outgoing ones. E.g. state $B$
would have the sum $\cancel{x_{\Tuple{B, \Char{c}, B}}} + x_{\Tuple{I_B,
\Char{b}, B}} = \cancel{x_{\Tuple{B, \Char{c}, B}}}  + x_{\Tuple{B, \Char{c},
F_B}} + x_{\Tuple{B, \Char{b}, I_B}}$ where we let $x_{\Tuple{q, l, c'}}$ refer
to the integer variable associated with a transition from state $q$ to state
$q'$ with label $l$. Note that the self-loop cancels itself out. This means that
for automata where a loop can become unreachable, these constraints do not fully
capture the constraints posed by the automaton, but requires additional
reachability constraints.

@@

% FIXME B-loopen inte nollad här! Uppdatera!

Several transitions are already marked as unuseable (have a transition variable
value of $0$). This includes any transition for the letter $\Char{d}$, as
$\CountOf{\Char{d}}(B) = \sum_{\Transition} 0 = 0$, since the register for
$\Char{d}$ is zero for all transitions in $B$.

Note that the self-loop on state $B$ is nonzero due to the above mentioned
self-balancing. However, a reachability computation will conclude that the state
is unreachable, and allow us to conclude that all its outgoing transitions must
be zero.

After removing all transitions concluded to be unuseable to produce new,
equivalent automata, we compute the product in \cref{fig:product}.

\begin{figure}[h]
    \centering 
    \includegraphics[scale=\autscale]{ab}
  \caption{An automaton encoding the constraint $\CountOf{\Char{a}}(B \times A')
  > \CountOf{\Char{b}}(B \times A')$.}\label{fig:product}
      %\Description[SHORT]{LONG}
\end{figure}

By putting off computing the product $B \times A'$ until after performing linear
reasoning on the number of times each transition must be taken and using that to
prune the terms, we have computed a smaller product than we would have with a
na\"ive approach. Moreover, since this automaton contains only one loop along
one main path, all runs through it can be described by the flow-preservation
equations described above without the need for costly connectivity-enforcing
constraints as in~\cite{generate-parikh-image}. We can now obtain a generalised
solution to our constraints by inspection.

>>>>

\begin{figure}[h]
    \centering 
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{a}
    \caption{The automaton $A'$, whose Parikh registers count the length of the
    string accepted by the automaton (register $l_A$), the start offset of the
    substring ($i$) and length ($n$) of the substring matching
    $\Regex{c\RegexOr{}dd}$. Note the symbolic transitions in the starting and
    accepting states matching any symbol in the alphabet!}\label{fig:aut_a}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{b}
    \caption{The automaton $B$, where the Parikh registers count the length of
    the string in register $l_B$.}\label{fig:aut_b}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The collection of automata we use as running
  examples}\label{fig:examples}
\end{figure}

In this section we will introduce an intuition for how a string constraint
problem in an automata-based solver like \OstrichPlus{} is translated to a
Parikh automata intersection problem and solved using \Calculus{}. First, by a
Parikh automaton we mean a standard non-deterministic finite automaton (NFA)
with a set of integer registers (Parikh registers) that are incremented (or
decremented) at each transition. In the automata of \cref{fig:examples}, we will
use the notation $a / \begin{bmatrix} x_+ \end{bmatrix}$ to mean that a
transition reads an input character $a$ and increments register $x$ by one. We
will omit zero-valued increments and assume that all registers are scoped to
their automaton. The increments are usually represented as a vector (hence the
brackets), but as it is mostly sparse here, we use the symbolic notation
$\begin{bmatrix} x_{1+} \end{bmatrix}$ rather than the more cumbersome $\left[
0, 1,  0 , \ldots, 0 \right]$.

Note that the labels of transitions can be ranges of characters with $\Sigma$
representing any character in the alphabet.

We use PCRE regular expression notation here and throughout the paper, writing
them $\Regex{\texttt{like this}}$. This means that $\RegexOr{}$ is alternation,
$\mathtt{*}$ the Kleene star, and $\mathtt{.}$, matches any single character.

For the length of a string $s$, we write $\Length{s}$.

\begin{example}\label{ex:string-constraints} Consider the following set of
    string constraints:
\begin{constraints}
    \item\label{const:s1-in-c-dd} $s_1 \in \Language(\Regex{c\RegexOr{}dd})$.
    \item\label{const:s2-in-b} $s_2 \in \Language(B)$, the language accepted by
    automaton~$B$ of \cref{fig:aut_b}.
    \item\label{const:s1-substring} $s_1 = \Substring(s_2, i, n)$, that is $s_1$ is an
    $n$-length substring of $s_2$ starting at offset $i$.
    \item\label{const:more-inside-than-before} $n > i$, that is what comes
    before $s_1$ in $s_2$ is shorter than $s_1$.
    \item\label{const:something-before-and-after} $i > 0 \land \Length{s_2} -i -n > 0$, that
    is there is at least one character in $s_2$ after $s_1$.
\end{constraints}
\end{example}

To solve \cref{ex:string-constraints} using \Calculus{} in a theorem proving
context, \cref{const:s1-in-c-dd,const:s2-in-b,const:s1-substring} can be
translated to the product of Parikh automata $B \times A'$, where $A'$
(\cref{fig:aut_a}) is the Parikh automata obtained by applying the encoding of
$\Substring$ described in \cite{ostrich-plus}. That description can then be used
to obtain values for the subsstring offsets $i, n$.

Intuitively, \cref{ex:string-constraints} are unsatisfiable. By inspecting the
automata, we realise that the path using $\Char{dd}$ in $A'$ is unusable since
are no $\Char{d}$-labelled transitions in $B$. This means that $s_1 = \Char{c}$,
and thus $n = 1$. That means that \cref{const:more-inside-than-before} implies
that $i = 0$. However, no path through $B$ passes by a $\Char{c}$ without a
preceding character. Therefore, already
\cref{const:s1-in-c-dd,const:s1-in-c-dd,const:s1-substring,const:more-inside-than-before}
together are unsatisfiable.

Our approach to the computation of the Parikh image (and therefore the solution
to the intersection problem posed by \cref{ex:string-constraints}) consists of
two applications of laziness; late computation of products of automata, and lazy
enforcement of connectivity constraints. Both are in contrast to the eager
approach to finding the Parikh image described in~\cite{generate-parikh-image},
that would have started by computing the product $A' \times B$, translated it to
a Presburger formula with $i, n$ etc as free variables, and then plugged in
\cref{const:more-inside-than-before,const:something-before-and-after}.

Starting with an interactive theorem prover and the goal of solving the Parikh
automata intersection problem, we associate each transition $\Transition$ of
$A'$ and $B$ with a fresh nonnegative integer variable from,
$\TransitionVar_\Transition$, identically to \cite{generate-parikh-image}. These
variables represent how many times each transition is taken. Hence, the final
counter value for counters $r_1, \ldots, r_k$ of each automaton is the
element-wise sum:
\begin{equation}\label{eq:counter-sums}
\begin{bmatrix} 
  r_1 \\
  \vdots\\
  r_k \\
\end{bmatrix} = \sum_\Transition \TransitionVar_\Transition \cdot 
  \IncrementVec_\Transition \text{ where $\IncrementVec_\Transition$ are the increments of transition $\Transition$}.
\end{equation}

@@

We proceed by adding linear constraints requiring transition variables to
represent a flow through their automaton by adding linear constraints stating
that the number of incoming transitions is equal to the number of outgoing ones.
E.g. state $B$ would have the sum $\cancel{x_{\Tuple{B, \Char{c}, B}}} +
x_{\Tuple{I_B, \Char{b}, B}} = \cancel{x_{\Tuple{B, \Char{c}, B}}}  +
x_{\Tuple{B, \Char{c}, F_B}} + x_{\Tuple{B, \Char{b}, I_B}}$ where we let
$x_{\Tuple{q, l, c'}}$ refer to the integer variable associated with a
transition from state $q$ to state $q'$ with label $l$. Note that the self-loop
cancels itself out. This means that for automata where a loop can become
unreachable, additional constraints would be required to ensure reachability.
However, being lazy, we put off that work until later.

@@

Several transitions are already marked as unuseable (have a transition variable
value of $0$) or always present (have a transition variable with a constant
value greater than zero) by just applying standard linear equation solving to
the flow equations described above. Note that this includes
\cref{const:more-inside-than-before,const:something-before-and-after}, which are
related to the transition variables through the sums described in
\cref{eq:counter-sums}. This means that we effectively can prune $A'$ of its
upper path when continuing the proof, as it is incongruent with the flow
equations and \cref{ex:string-constraints}. Note that the per-automata length
counting registers have been assigned the same solver variable $l_A$. Since they
have to be equal, either one can be used in both automata.

Notably, one such conclusion is that $i > 0$ from
\cref{const:something-before-and-after}. From flow preservation we have that the
$F_{A'}$ to $F_{A'}$ transition is represented by $l_A - n - i$.
\footnote{Intuitively, the total length is made up of what is before, inside and
after the substring.} This in turn through \cref{const:more-inside-than-before}
forces $n \geq 2$, which in turn (through flow-preservation) forces a path from
$I_{A'}$ to $F_{A'}$ through $D$, rather than the single-transition path from
$I_{A'}$ to $F_{A'}$. This immediately implies that the problem is
unsatisfiable, since we know this transition has no correspondance in automaton
$B$.

However, \Calculus{} does not at this point, and we press on with concluding
that none of the loops of $A'$ at this stage can become disconnected since
they are at the initial and final states, both of which are trivially reachable
for any assignments of the non-constant transition variables. This means that
all valid counter assignments of $A'$ are described by \cref{eq:counter-sums}
and the flow equations alone without the need for expensive connectivity
constraints.

After concluding this, we have a choice of two paths through automaton $B$; the
upper through state $A$ or the lower through state $B$. We case split on a
problem-aware basis by selecting a transition variable that would disconnect
some strongly connected component of automaton $B$, in this case the transition
form $I_B$ to state $B$. We divide it into the cases used ($> 0$) or unused ($ =
0$) since those are the threshold values of the calculus, preferring the latter
option on the first-fail principle.

At this point we have two small, stick-like automata after discounting all
transitions with a corresponding transition variable that is zero, and can try
to compute their product. Doing so, we will immediately notice that the
$\Char{d}$-labelled transition of automaton $A'$ has no correspondence in $B$,
leading to an empty product. We close the proof goal and backtrack with the
opposite constraint, choosing the upper $I_B$ to $A$ transition this time, the
only potentially productive choice we have left. However, proceding with
calculation of the product from the almost identical terms will be similarly
improductive, and we must conclude that the problem is unsatisfiable.

By putting off computing the product $B \times A'$ until after performing linear
reasoning on the number of times each transition must be taken and using that to
prune the terms, we have computed a smaller product than we would have with a
na\"ive approach. Moreover, since this automaton contains only one loop along
one main path, all runs through it can be described by the flow-preservation
equations described above without the need for costly connectivity-enforcing
constraints as in~\cite{generate-parikh-image}.

### 2023-06-12 introduction.tex
Use a new, unsatisfiable example
This version uses a new example which is hopefully easier to follow. The linear algebra is slightly breakneck, but there is little I can do about that.

This section is now, hopefully, less jargon-heavy!

Closes: #7, #31, #26, #30,  #29, #28.

Parikh image computations as well as Parikh automata feature extensively in
string solvers, including as mentioned above \Ostrich{} and \OstrichPlus{}
\cite{ostrich,ostrich-plus}, but also forms the basis of Trau~\cite{trau-pldi},
and occurs in \textsc{Sloth}~\cite{sloth}. Parikh images frequently appear when
introducing cardinality constraints like length or string indexing. The
state-of-the-art approach to handling Parikh image computation is to
over-approximate the Parikh image of a product of $k$~automata
$\ParikhMap(\Language(\Automaton_1) \cap \ldots \cap \Language(\Automaton_k))$
with the conjunction of the automata's parikh maps, $\bigwedge_{i=1}^{k}
\ParikhMap(\Automaton_i)$. This approach works only for unsatisfiable instances,
and will require falling back to computing the product of the automata before
using the standard approach for finding its image originally presented
in~\cite{generate-parikh-image}.

>>>>

Parikh image computations as well as Parikh automata feature extensively in
string solvers, including as mentioned above \Ostrich{} and \OstrichPlus{}
\cite{ostrich,ostrich-plus}, but also forms the basis of Trau~\cite{trau-pldi},
and occurs in \textsc{Sloth}~\cite{sloth}. Parikh images frequently appear when
introducing cardinality constraints like length or string indexing. The
state-of-the-art approach to handling Parikh image computation is to
over-approximate the Parikh image of a product of $k$~automata
$\ParikhMap(\Language(\Automaton_1) \cap \ldots \cap \Language(\Automaton_k))$
with the conjunction of the automata's Parikh maps, $\bigwedge_{i=1}^{k}
\ParikhMap(\Automaton_i)$. This approach works only for unsatisfiable instances,
and will require falling back to computing the product of the automata before
using the standard approach for finding its image originally presented
in~\cite{generate-parikh-image}.
